---
title: "Case-control dataset and descriptive analysis"
author: "Noemi Nicole Piga"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    code_folding: "hide"
    toc: true
    toc_float: true
    theme: united
---


```{r, setup, hide = TRUE}
#library reticulate to be able to add chuncks in another language
library(reticulate)
```

```{bash, hide = TRUE, eval = FALSE}
# To run this .rmd file in a terminal being in the project folder:
#module unload R/4.2.1
#module load R/4.1.0
#export PATH=${PATH}:/cm/shared/apps/R/deps/rstudio/bin/pandoc
#file="./report/Case_control_descriptive.Rmd"
#module unload R/4.2.1
#module load R/4.1.0
#Rscript -e 'rmarkdown::render("'$file'")'
```
# Rationale
Create the case and control dataset to run the genome-wdie association analysis. Descriptive analysis of several demographical and biological features.

# Analysis
Code: SA_UKBiobank_datafields_create_MasterTable.sh
```{bash, code = readLines('/home/n/nnp5/PhD/PhD_project/UKBiobank_datafields/scripts/SA_UKBiobank_datafields_create_MasterTable.sh'), eval=FALSE}
```
Code: SA_UKBiobank_datafields_create_MasterTable.py
```{bash, code = readLines('/home/n/nnp5/PhD/PhD_project/UKBiobank_datafields/scripts/SA_UKBiobank_datafields_create_MasterTable.py'), eval=FALSE}
```
Code: Add_genAnc_PCs_demo.r
```{bash, code = readLines('/home/n/nnp5/PhD/PhD_project/UKBiobank_datafields/scripts/Add_genAnc_PCs_demo.r'), eval=FALSE}
```
Code: merge_age_onset.r
```{bash, code = readLines('/home/n/nnp5/PhD/PhD_project/UKBiobank_datafields/scripts/merge_age_onset.r'), eval=FALSE}
```
Code: control_definition.sh
```{bash, code = readLines('/home/n/nnp5/PhD/PhD_project/UKBiobank_datafields/scripts/control_definition.sh'), eval=FALSE}
```
Code: control_definition_V2.sh
```{bash, code = readLines('/home/n/nnp5/PhD/PhD_project/UKBiobank_datafields/scripts/control_definition_V2.sh'), eval=FALSE}
```
Code: pheno_cov_file.r
```{bash, code = readLines('/home/n/nnp5/PhD/PhD_project/REGENIE_assoc/src/pheno_cov_file.r'), eval=FALSE}
```
Code: case_control_descriptive.R
```{bash, code = readLines('/home/n/nnp5/PhD/PhD_project/REGENIE_assoc/src/case_control_descriptive.R'), eval=FALSE}
```
Cases were defined as having either:


 *all-comer asthma with medications within British Thoracic Society 2019 guidelines step 4-5 filter (see 'Phenotype Definition');


 *hospitalisation due to asthma as primary reason (Data-Field 41234, value ‘L1’);
 
 
 *death due to asthma as primary reason (Data-Fields 40001, ICD-10 code value ‘J45’ or ‘J46’);
 
 
 *European ancestry
 
<br>
Controls were defined as individuals without the following (Figure 6):


 *Either asthma, emphysema, chronic bronchitis or COPD diagnostic codes (Table S8);


 *Self-reported asthma or any other major respiratory condition (Data-field 20002, value ‘1111’ for asthma and ‘1072/1466/1130/1131/1132/1133/1134/1135/1660/1139/1140/1141/1142/1143/1144/1467/1164/1168/1534/1559/1560/1561/1562/1563/1465’ for other respiratory conditions);
 
 
 *FEV1/FVC ratio <= 70% (Data-Field 20150, 20151);
 
 
 *Predicted percentage FEV1 < 60% (Data-Field 20154);
 
 
 *General practice prescriptions or self-reported medications for asthma (Data-Field 42039, 20003);
 
 
 *Absence of GP prescription or self-reported medication records (Data-Field 42039, 20003);


 *Non-European ancestry

I selected five age- and sex-matched controls for every case.
<br>
In order to investigate my case-control cohort, I compared demographic information including genetic sex (Data-Field 22001), age at recruitment (Data-Field 21022), and smoking status (Data-Field 20116) as defined in U-BIOPRED63 using pack years of smoking (Data-Field 20616). I also compared clinical traits: age at onset of asthma (merged Data-Field 3786 and 22147), BMI (Data-Field 23104), hospitalisation due to asthma (Data-Field 41234, ICD-9 and ICD-10, ‘level 1’), eosinophil count (Data-Field 30150), neutrophil count (Data-Field 30140), percentage predicted FEV1 (in house pre-processed from Data-Field 20154), FEV1/FVC ratio (in-house pre-processed from Data-Field 20150, 20151), prednisolone use (from prescription records, Data-field 42039).

# Results
I obtained 4,021 individuals with hospitalisation and 45 individuals with death primarily due to asthma. I merged this result with the European individuals within British Thoracic Society (BTS) stage 4-5 obtaining a final number of 7,681 cases. I selected five age- and sex-matched controls for every case, obtaining 38,405 European controls.
<br>
Overall, the mean age is 57 years old. The dataset has a higher proportion of females compared to males (63.95% vs 36.05%). Cases exhibit higher BMI, lower lung function, and higher blood cell counts of eosinophils and neutrophils compared to controls. There are a more smokers for cases than for controls. Amongst cases, 61.82% self-report adult-onset asthma, 8.37% have been hospitalised with asthma as primary cause at least once, and 66.70% have been prescribed for prednisolone at least once (see 'Probation' Table1).


```{r, setup, hide = TRUE}
library(kableExtra)
library(readxl)
demo <- read_xlsx("/home/n/nnp5/PhD/PhD_project/REGENIE_assoc/data/Descriptive_table_EUR.xlsx", sheet="Descriptive_broadsevereasthma")
colnames(demo)[1] <-"Traits"
colnames(demo)[2] <- "Sub-traits"
demo <- demo[-c(1,3,7),]
#https://cran.r-project.org/web/packages/kableExtra/vignettes/awesome_table_in_html.html#Grouped_Columns__Rows help here
kbl(demo) %>% 
  kable_paper("striped", full_width = F) %>%
  pack_rows("Age, years", 1,10) %>%
  pack_rows("Sex", 2,3)
  

```

# Conclusion